---
title: "R Notebook"
output: html_notebook
---


##Use a scatterplot to explore the relationship between GDP per capita and life expectancy
##Bonus question: Can you map the population to a particular aesthetic so that it can be displayed on the chart too?
```{r}
gapminder
library(gapminder); library(tidyverse); library(scales); library(dplyr)
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point()

```
```{r}
ggplot(gapminder, aes (x = gdpPercap, y = lifeExp, size = pop)) +
  geom_point()
```
##Line Chart
```{r}
gapminder
library(gapminder); library(tidyverse); library(scales); library(dplyr)
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_line()
```


##Draw a line chart showing the trend for average gdp per capita through time for each continent
```{r}
gapminder %>%
group_by(year, continent) %>%
summarize(meangdpPercap = mean(gdpPercap), na.rm = T) %>%
  ggplot(data = ., aes(x = year, y = meangdpPercap)) +
  geom_line(aes(color=continent))
```
##Draw a histogram to show the characteristics of population
```{r}
ggplot(gapminder, aes(x = pop)) +
  geom_histogram() +
scale_x_log10()
```
##Draw a Bar chart
```{r}
ggplot(gapminder, aes(x = year, weight = pop)) +
  geom_bar() +
scale_x_log10()
```
##Draw a bar chart showing the average lifeExp across years
```{r}
by_year <- gapminder %>%
group_by(year) %>%
summarize(meanLifeExp = mean(lifeExp), na.rm = T) %>%
  ggplot(aes(x = year, y = meanlifeExp, by_year)) +
  geom_bar(aes(color = year))
```




