---
title: "Numbers Ninja Revision 1"
author: |
  | Hari Subhash
  | Data Scientist @NRGI
date: "`r Sys.Date()`"
output:
  html_notebook:
    highlight: kate
    smart: yes
    theme: cosmo
  html_document:
    df_print: paged
    includes: assets/header.html
css: assets/custom.css
---

`extractoR`

##Todo

1. What is the class of function?
2. Share data camp join video

##Concepts to cover

1. ~Data types~
2. ~Data structures~
3. ~What is your environment?~
4. ~What are functions?~
5. ~What are packages?~
6. ~What is declaritive code? The opposite is called imperative.~
7. ~What is composition? Every function should do only one thing~
8. ~Review of dplyr (focus on group_by() + summarise())~
9. Review of ggplot
10. ~Review of data joins~

End of next week you should have enough skills to get started


##Data Types

1. characters
2. numeric
3. integer
4. logical/boolean

```{r}
library(tidyverse); library(nycflights13)
ninjaNames <- c("Anne", "Alonso","Hari", "Solongo")
ninjaRunMiles <- c(26L, 3L, 10L)


if (typeof(ninjaNames) == "numeric") {
    ##do something
    print("hello world")
} else {
    print("wrong class")
}

paste(ninjaNames, c("awesome", "great"))

typeCoercedVector <- c(ninjaNames, ninjaRunMiles)
typeCoercedVector
typeof(typeCoercedVector)
```

```{r}

```

##Data Structures

1. Data Frames/tibbles
2. Lists
3. Matrices
4. Arrays
5. Vector

Data frames

```{r}
dim(flights)
```

List
```{r}
testList <- list(flights = flights, 
                 ninjaNames = ninjaNames, 
                 ninjaRunMiles = ninjaRunMiles)

testList$flights
```

```{r}
testList[[1]]
```




##What is your environment?

```{r}
ninjaNames
```

```{r}
flights %>% 
    filter(dep_delay > 100)
```


function in dplyr that we have learnt: ` %>% `, ``


```{r}
class(print)
```

D.R.Y: Don't Repeat Yourself. Functions are extremely useful for that.


```{r}

greetNamesFunc <- function(inputNames) {
    
    if (typeof(inputNames) == "character") {
        for (i in 1:length(inputNames)) {
            print(paste("Hi", inputNames[i], "how are you?"))
        }    
    } else {
        print("Requires a character vector")
    }
}

newCharVector <- c("blaj", "la")

greetNamesFunc(inputNames = ninjaNames)
greetNamesFunc(ninjaRunMiles)
greetNamesFunc(newCharVector)
```


##dplyr

group_by
```{r}
flights %>% 
    select(origin, dep_delay, arr_delay) %>% 
    group_by(origin, dep_delay) %>% 
    summarise(meanDepDelay = mean(dep_delay, na.rm = T))
```



##data joins
primary key
```{r}
band_members
```


```{r}
ninja_instruments <- bind_rows(band_instruments, tibble(name = "John", plays = "keyboards"))
ninja_instruments$alive <- c(FALSE, TRUE, TRUE, FALSE)
ninja_instruments
```

foreign key

```{r}
full_join(band_members, ninja_instruments, by = "name")
```



```{r}
flights
```
```{r}
names(flights)
```


```{r}
left_join(flights, airlines, by = "carrier")
```


```{r}
weather
```


This plot is a basic scatter between visibility and departure delay. 
```{r}
plotData <- flights %>% 
    select(origin, month, day, hour, dep_delay) %>% 
    left_join(., weather, by = c("origin", "month", "day", "hour")) %>% 
    filter(precip != 0) 


ggplot(plotData, aes(x = visib, y = dep_delay)) + 
    geom_point()
```

Lets try and group the data by visibility and see what happens. Did not work. Lets try something else.
```{r}
plotData %>% 
    group_by(visib) %>% 
    summarise(meanDD = mean(dep_delay, na.rm = T)) %>% 
    ggplot(., aes(x = visib, y = meanDD)) +
    geom_point()
```

