---
title: "R Notebook"
output: html_notebook
---


##Use a scatterplot to explore the relationship between GDP per capita and life expectancy
##Bonus question: Can you map the population to a particular aesthetic so that it can be displayed on the chart too?
```{r}
library(gapminder); library(tidyverse); library(scales); library(dplyr)

gapminder
```

```{r}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point()

```


```{r}
ggplot(gapminder, aes (x = gdpPercap, y = lifeExp, size = pop)) +
  geom_point()
```

##Line Chart

```{r}
gapminder %>% 
    group_by(year, continent) %>% 
    summarise(meanLifeExp = mean(lifeExp, na.rm = T), meanGDPperCap = mean(gdpPercap, na.rm = T)) %>% 
    ggplot(., aes(x = meanGDPperCap, y = meanLifeExp)) +
    geom_line() + 
    facet_wrap(~continent)
```


##Draw a line chart showing the trend for average gdp per capita through time for each continent
```{r}
gapminder %>%
group_by(year, continent) %>%
summarize(meangdpPercap = mean(gdpPercap), na.rm = T) %>%
  ggplot(data = ., aes(x = year, y = meangdpPercap)) +
  geom_line(aes(color = continent))
```

##Draw a histogram to show the characteristics of population
```{r}
gapminder %>% 
    filter(year == 2007) %>% 
    ggplot(., aes(x = pop)) +
    geom_histogram(bins = 200)
```

##Draw a Bar chart
```{r}
gapminder %>% 
    group_by(year) %>% 
    summarise(meanPop = mean(pop, na.rm = T)) %>% 
    ggplot(., aes(x = year, y = meanPop)) +
    geom_col()
```

##Draw a bar chart showing the average lifeExp across years
```{r}
gapminder %>%
    group_by(year) %>%
    summarize(meanLifeExp = mean(lifeExp, na.rm = T)) %>%
    ggplot(data = ., mapping = aes(x = year, y = meanLifeExp)) +
    geom_col(width = 2)
```




